---
title: "Peak Bloom Prediction Demo"
author: "Eager Learner"
date: "01/05/2024"
lang: en-US
format:
  html:
    embed-resources: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, 
                      message = FALSE,
                      fig.align = 'center',
                      out.width = '80%')
```

## Instructions

In this analysis we demonstrate a simple way of predicting the peak bloom date over the next decade for all five locations required by the competition.
The models are simple in that they only use the past bloom dates observed at each location---no other covariates or additional information are considered.
At the end of this document ([Appendix A](#appendix-rnoaa)), we demonstrate a simple way to get historic temperature data for the four locations via the [NOAA web API](https://www.ncdc.noaa.gov/cdo-web/).

For this demo analysis we are using methods from the _tidyverse_ of R packages.
They can be installed via

```{r}
#| eval: false
install.packages('tidyverse')
```

and then loaded via

```{r}
library(tidyverse)
```


## Loading the data

The data for the five sites is provided as a simple text file in CSV format.
Each file contains the dates of the peak bloom of the cherry trees at the respective site, alongside the geographical location of the site.

The six columns in each data file are

* _location_ a human-readable location identifier (`string`).
* _lat_ (approximate) latitude of the cherry trees (`double`).
* _long_ (approximate) longitude of the cherry trees (`double`).
* _alt_ (approximate) altitude of the cherry trees (`double`).
* _year_ year of the observation (`integer`).
* *bloom_date* date of peak bloom of the cherry trees (ISO 8601 date `string`). The "peak bloom date" may be defined differently for different sites
* *bloom_doy* days since January 1st of the year until peak bloom (`integer`). January 1st corresponds to `1`.

In R, the data files can be read with `read.csv()` and concatenated with the `bind_rows()` function:

```{r}
cherry <- read.csv("data/washingtondc.csv") |> 
  bind_rows(read.csv("data/liestal.csv")) |> 
  bind_rows(read.csv("data/kyoto.csv")) |> 
  bind_rows(read.csv("data/vancouver.csv")) |> 
  bind_rows(read.csv("data/nyc.csv"))
```

For example, the latest 3 observations for each location can be extracted with:

```{r}
cherry |> 
  group_by(location) |> 
  slice_tail(n = 3)
```

## Visualizing the time series


```{r}
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Time series of peak bloom of cherry trees since 1880 at four different sites.
cherry |> 
  filter(year >= 1880) |>
  ggplot(aes(x = year, y = bloom_doy)) +
  geom_point() +
  geom_step(linetype = 'dotted', color = 'gray50') +
  scale_x_continuous(breaks = seq(1880, 2020, by = 20)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = "Year", y = "Peak bloom (days since Jan 1st)")
```

## Predicting the peak bloom

A simple method to predict peak bloom date in the future is to fit a least-squares line through the observed dates and extrapolate the regression function.
We want to have a separate line for each location, hence we tell R to estimate _interaction_ effects.
We only use data from 1880 to fit the trends, as prior data may not be as reliable/relevant.
Since for New York City we only have one observed bloom date, which is not enough to fit a linear regression model, we will omit the site from this simple analysis.

```{r}
# Fit simple least-squares lines for all sites.
cherry_no_nyc <- cherry |> 
  filter(location != 'newyorkcity')
ls_fit <- lm(bloom_doy ~ 0 + location + location : year, 
             data = cherry_no_nyc, subset = year >= 1880)
```

This simple linear regression functions suggest a trend toward earlier peak bloom at the sites.
We can compute the actual predictions using the `predict()` function and 

```{r}
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from simple linear regression models fitted to
#|   four sites.
# Compute the predictions for all 4 sites
cherry_no_nyc_grid <- expand_grid(location = unique(cherry_no_nyc$location),
                                  year = 1880:2025)
predictions <- cherry_no_nyc_grid |> 
  bind_cols(predict(ls_fit, newdata = cherry_no_nyc_grid, 
                    interval = 'prediction', level = 0.9)) |> 
  rename(prediction = fit, lower = lwr, upper = upr)

# Plot the predictions alongside the actual observations for 2015 up to 2023.
cherry |> 
  right_join(predictions, by = c('year', 'location')) |>
  filter((location == 'vancouver' & year >= 2021) |
           (location != 'vancouver' & year >= 2015)) |>
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = 'black', linetype = '22', linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = c(2015, 2018, 2021, 2024)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = "Year", y = "Peak bloom (days since Jan 1st)")
```

Based on this very simple model, the peak bloom dates at these sites are:

```{r}
#' Small helper function to convert the day of year to
#' the actual date.
#' 
#' @param year year as an integer
#' @param doy day of the year as integer (1 means January 1st)
#' @return date string
doy_to_date <- function (year, doy) {
  strptime(paste(year, doy, sep = '-'), '%Y-%j') |> # create date object
    strftime('%Y-%m-%d') # translate back to date string in ISO 8601 format
}

predictions |> 
  filter(year == 2025) |> 
  mutate(prediction = round(prediction),
         lower = floor(lower),
         upper = ceiling(upper),
         prediction_date = doy_to_date(year, prediction))
```

## Extrapolating to Vancouver, BC and New York City

For the cherry trees in Vancouver, BC and New York City few historical observations are available.
This shows in the simple analysis above in the very wide prediction interval.
The trees in Vancouver, BC are located approximately at 49.2236916°N (latitude), -123.1636251°E (longitude), 24 meters above sea levels (altitude).
Casual observations for Vancouver, BC have been recorded in the way of photos posted to the [VCBF Neighbourhood Blog for Kerrisdale](https://forums.botanicalgarden.ubc.ca/threads/kerrisdale.36008/).
You can search the forum for the keywords "Akebono" (i.e., the name of the cultivar) and "Maple Grove Park" (i.e., the location of the trees).

We need to *extrapolate* from what we have learned about the peak bloom dates in the other locations to Vancouver and NYC.
The simple model we have fitted above, however, does not allow us to transfer any knowledge from the other sites -- we have only used the history trend at the respective sites.

Although the climate in Vancouver and NYC is different from the other locations, the simplest way to borrow information from the other locations is to average across these three sites.
Hence, we want to fit a straight line through the peak bloom dates, ignoring the actual site:

```{r}
# Fit simple least-squares lines for all sites.
# We use larger weights for observations from Vancouver than for the other sites
ls_fit_for_van <- lm(bloom_doy ~ year, data = cherry, subset = year >= 1880,
                     weights = (location == 'vancouver') +
                       0.2 * (location != 'vancouver'))

vancouver_grid <- tibble(location = 'vancouver', year = 2023:2025)
predictions_vancouver <- vancouver_grid |> 
  bind_cols(predict(ls_fit_for_van, newdata = vancouver_grid,
                    interval = 'prediction', level = 0.9)) |> 
  rename(prediction = fit, lower = lwr, upper = upr)
```

Not surprisingly, the predicted peak bloom date for Vancouver and NYC is now very similar to the other 3 sites:
```{r}
predictions_vancouver
```

We can check the predictions against the data from previous competition years.
```{r}
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from a simple linear regression model for
#|   Vancouver using data from all four sites.
# Plot the predictions alongside the actual observations for 2015 up to 2023.
cherry |> 
  right_join(predictions_vancouver, by = c('year', 'location')) |> 
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = 'black', linetype = '22', linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = 2022:2024) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = "Year", y = "Peak bloom (days since Jan 1st)")
```

If satisfied with the predictions, we can use them instead of the predictions from before.
```{r}
predictions <- predictions |> 
  filter(location != 'vancouver') |> 
  bind_rows(predictions_vancouver)
```

## Extrapolating to New York City, NY using USA-NPN data

Similar to Vancouver, BC, only few historical observations are available for our location in New York City, NY.
There are some historical observations dating back to 2019 in the data provided by USA-NPN.
The Washington Square Park has site id 32789 and the Yoshino cherry you should predict has species id 228.

```{r}
#| label: load-historical-data-nyc
nyc_data_npn <- read_csv("data/USA-NPN_status_intensity_observations_data.csv") |> 
  filter(Site_ID == 32789,
         Species_ID == 228) |> 
  mutate(Observation_Date = as_date(Observation_Date, format = '%m/%d/%y'))
```

This data, however, needs to be transformed as it only contains individual observations of the phenophase and not the actual peak bloom date.
For simplicity, we take the first day someone observed the flowers to be open as the peak bloom day.
This could be done in a more sophisticated way by also looking at the reported intensity value.

```{r}
#| warning: true
nyc_data <- nyc_data_npn |> 
  arrange(Observation_Date) |> 
  mutate(year = year(Observation_Date)) |> 
  group_by(year) |> 
  summarize(first_flower_index = min(which(Phenophase_Status == 1)),
            bloom_date = strftime(Observation_Date[first_flower_index], format = '%Y-%m-%d'),
            bloom_doy = Day_of_Year[first_flower_index],
            .groups = 'drop') |> 
  filter(!is.na(bloom_doy)) |> 
  select(-first_flower_index) |> 
  mutate(location = 'newyorkcity')

cherry_with_nyc <- cherry |> 
  bind_rows(nyc_data)
```

For 2020, no bloom was reported, hence the warning.

Using the same steps as for Vancouver, BC, a simple linear model can be fitted.
```{r}
# Fit simple least-squares lines for all sites.
# We use larger weights for observations from NYC than for the other sites
ls_fit_for_nyc <- lm(bloom_doy ~ year, data = cherry_with_nyc, subset = year >= 1880,
                     weights = (location == 'newyorkcity') +
                       0.2 * (location != 'newyorkcity'))

nyc_grid <- tibble(location = 'newyorkcity', year = 2022:2025)
predictions_nyc <- nyc_grid |> 
  bind_cols(predict(ls_fit_for_nyc, newdata = nyc_grid,
                    interval = 'prediction', level = 0.9)) |> 
  rename(prediction = fit, lower = lwr, upper = upr)

predictions_nyc
```

We can check the predictions against the data from previous competition years.
```{r}
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from a simple linear regression model for
#|   Washington Square Park in NYC using data from all five sites.
# Plot the predictions alongside the actual observations for 2015 up to 2023.
cherry_with_nyc |> 
  right_join(predictions_nyc, by = c('year', 'location')) |>
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = 'black', linetype = '22', linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = 2022:2025) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = "Year", y = "Peak bloom (days since Jan 1st)")
```

If satisfied with the predictions, we can use them instead of the predictions from before.
```{r}
predictions <- predictions |> 
  filter(location != 'newyorkcity') |> 
  bind_rows(predictions_nyc)
```

## Submission

To submit your entries, enter the predicted bloom dates from your models in the submission form.

```{r}
predictions |> 
  filter(year == 2025) |> 
  mutate(predicted_date = strptime(paste(year, prediction), "%Y %j") |> 
           as_date())
```



## Appendix: Adding Covariates {#appendix-rnoaa}

We encourage you to find additional publicly-available data that will improve your predictions.
For example, one source of global meteorological data comes from the Global Historical Climatology Network (GHCN), available through the NOAA web API.

To use the web API, you first need a web service token.
You can request this token (free of charge) via <https://www.ncdc.noaa.gov/cdo-web/token>.
Once you have been issued the token, note it somewhere in your code (or make it available through an environment variable):

```{r}
#| eval: false
NOAA_WEB_API_TOKEN <- '...'
# or
NOAA_WEB_API_TOKEN <- Sys.getenv("NOAA_WEB_API_TOKEN")
```

```{r}
#| echo: false
NOAA_WEB_API_TOKEN <- Sys.getenv("NOAA_WEB_API_TOKEN")
```


To connect to and use the web API you may use the following R packages:

```{r}
#| eval: false
install.packages("httr2")
install.packages("jsonlite")
```

and the loaded via

```{r}
library(httr2)
library(jsonlite)
```

The stations closest to the sites for the competition with continuously collected maximum temperatures are USC00186350 (Washington D.C.), GME00127786 (Liestal), JA000047759 (Kyoto), CA001108395 (Vancouver) and 

```{r}
NOAA_API_BASE_URL <- "https://www.ncei.noaa.gov/cdo-web/api/v2/data"

# Define the station IDs for the specified locations
stations <- c(
  "washingtondc" = "GHCND:USW00013743",
  "vancouver"    = "GHCND:CA001108395",
  "newyorkcity"  = "GHCND:USW00014732",
  "liestal"      = "GHCND:SZ000001940",
  "kyoto"        = "GHCND:JA000047759")
```

As a simple demonstration, we retrieve the average seasonal maximum daily temperature (in 1/10 °C) from these stations using our own `get_temperature()` function, which wraps the `ghcnd_search()` function in the `rnoaa` package. (N.b. `ghcnd_search()` returns a list. Each element of the list corresponds to an element of the `var` argument.)

```{r}
nested_to_tibble <- function (x) {
  # Determine the variable names in the response
  variable_names <- map(x, names) |> 
    unlist(use.names = FALSE) |> 
    unique()
  
  names(variable_names) <- variable_names

  # Reshape the response from a nested list into a table
  map(variable_names, \(i) {
    map(x, \(y) {
      if (is.null(y[[i]])) {
        NA_character_
      } else {
        y[[i]]
      }
    }) |> 
      unlist(use.names = FALSE)
  }) |> 
    as_tibble()
}

get_daily_avg_temp <- function(station_id, start_date, end_date,
                               api_key, base_url, window_size = 300) {
  windows <- seq(as_date(start_date),
                 as_date(end_date) + days(window_size + 1),
                 by = sprintf("%d days", window_size))
  
  batches <- map2(windows[-length(windows)], windows[-1] - days(1), \(from, to) {
    if (from > Sys.Date()) {
      return(NULL)
    }
    response <- tryCatch(
      request(base_url) |> 
        req_headers(token = api_key) |> 
        req_url_query(
          datasetid = "GHCND",
          stationid = station_id,
          datatypeid = "TAVG,TMAX",
          startdate = from,
          enddate = min(as_date(to), Sys.Date()),
          units = "metric",
          limit = 1000
        ) |> 
        req_retry(max_tries = 10) |> 
        req_perform() |> 
        resp_body_json(),
      
      httr2_http = \(cnd) {
        rlang::warn(sprintf("Failed to retrieve data for station %s in time window %s--%s",
                            station_id, from, to),
                    parent = cnd)
        NULL
      })
  })
  
  map(batches, \(x) nested_to_tibble(x$results)) |> 
    list_rbind() |> 
    mutate(date = as_date(date))
}
```

```{r}
#| cache: true
historic_temperatures <- cherry |> 
  group_by(location) |> 
  summarize(start_date = sprintf('%d-01-01', pmax(1970, min(year)) - 1)) |> 
  left_join(tibble(location = names(stations),
                   station_id = stations),
            by = 'location') |> 
  group_by(location) |> 
  group_modify(\(x, gr) {
    get_daily_avg_temp(station_id = x$station_id,
                       start_date = x$start_date,
                       end_date = Sys.Date(),
                       api_key = NOAA_WEB_API_TOKEN,
                       base_url = NOAA_API_BASE_URL)
  })
```

```{r}
historic_temperatures |>
  filter(datatype == 'TMAX') |> 
  ggplot(aes(x = date, y = value)) + 
  geom_line() +
  labs(x = "Year", y = "Average maximum temperature (°C)") +
  facet_grid(rows = vars(location))
```

A simple model may simply take the average maximum winter temperature (Dec. 1st of the previous year until end of February) into account:

```{r}
avg_winter_temp <- historic_temperatures |> 
  filter(datatype == 'TMAX') |> 
  mutate(year = case_when(
    month(date) < 3 ~ year(date),
    month(date) == 12 ~ year(date) + 1,
    TRUE ~ NA_integer_
  )) |> 
  filter(!is.na(year), year >= 1970) |> 
  group_by(location, year) |> 
  summarize(avg_tmax = mean(value),
            .groups = 'drop')
```

```{r}
avg_winter_temp |>
  ggplot(aes(x = year, y = avg_tmax)) + 
  geom_line() +
  labs(x = "Year", y = "Average maximum temperature (°C)") +
  facet_grid(rows = vars(location))
```

Using these average temperature, we can predict the peak bloom date again using linear regression with location-specific temporal trends and a global effect of average winter temperatures. 

```{r}
ls_fit_with_temp <- cherry |>
  inner_join(avg_winter_temp,
             by = c("location", "year")) |>
  lm(formula = bloom_doy ~ year * location + avg_tmax)

cherry_grid <- expand_grid(location = unique(cherry$location),
                           year = 1990:2025) |> 
  inner_join(avg_winter_temp,
             by = c("location", "year"))

predictions_from_temp <- cherry_grid |> 
  mutate(pred_bloom = predict(ls_fit_with_temp, newdata = cherry_grid))

predictions_from_temp |> 
  left_join(cherry,
             by = c("location", "year")) |> 
  ggplot(aes(x = year)) +
  geom_point(aes(y = bloom_doy)) +
  geom_line(aes(y = pred_bloom)) +
  facet_grid(rows = vars(location))
```

The following plot shows a comparison of predictions for Vancouver using the two methods described in this demo.

```{r}
predictions_vancouver |>
  left_join(predictions_from_temp,
            by = c("location", "year")) |> 
  select(year, pred_temporal = prediction, pred_temp = pred_bloom) |> 
  pivot_longer(cols = -year) |> 
  mutate(name = if_else(name == "pred_temporal", 
                        "Method 1: location-based model", 
                        "Method 2: temperature-based model")) |>
  ggplot() +
  aes(x = year, y = value, linetype = name) +
  geom_line() +
  scale_x_continuous(breaks = 2022:2024) +
  labs(x = "Year", linetype = "",
       y = "Predicted peak bloom (days since Jan 1st) for Vancouver") +
  theme(legend.position = "bottom")
```


